Characterizing subarctic European vegetation for land surface modeling of ozone pollution and climate risk

### 2021-09-22 VegOzone.tex
Adjust acknowledgment
Update from Overleaf

%\title{The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate}
\title{Characterizing subarctic biomes for land surface modeling of pollution and climate risk}

@@

\Author[1]{Stefanie}{Falk}
\Author[2]{Ane Victoria}{Vollsnes}
\Author[2]{Aud}{Eriksen}
\Author[3]{Lisa}{Emberson}
\Author[3]{Connie}{O'Neill}
\Author[1]{Frode}{Stordal}
\Author[1]{Terje}{Koren Berntsen}

\affil[1]{Department of Geosciences, University of Oslo, Oslo, Norway}
\affil[2]{Department of Biosciences, University of Oslo, Oslo, Norway}
\affil[3]{Department of Environment and Geography, University of York, UK}

@@

\begin{abstract}
  The unique vegetation biomes of the subarctic region acclimatized to extremes of cold and midnight sun are likely to be at threat from the combined impacts of climate change and increasing ozone (\chem{O_3}) concentrations. The atmospheric and climatic characteristics of the subarctic are known to lead to pronounced peak \chem{O_3} concentrations in spring. To date, only a few studies have been conducted to assess the response of subarctic vegetation to variations in climate and air pollution, at least by comparison with other parts of Europe. This study looks to fill this knowledge gap by examining trends in climate variables and \chem{O_3} concentrations that have occurred over the past few decades. These climatological and \chem{O_3} data are used to assess the extent to which two recent years (2018 and 2019) deviate from climatic and \chem{O_3} concentration norms and how these current (and potentially more frequent future) deviations may influence \chem{O_3} damage to subarctic vegetation. We find that 2018 was an anomalous warm and bright year (in particular in spring and early summer) with higher than average \chem{O_3} concentrations in April/May and a high number of episodic peak \chem{O_3} concentration above $40\,\unit{ppb}$ in June--August, the latter in part attributable to an increase in forest fires in the Northern Hemisphere as well as warmer and sunnier conditions. We apply a flux-metric to determine \chem{O_3} risk and damage to vegetation as a function of \chem{O_3} concentration, climate variables, and species-specific physiology. We find that bespoke parameterizations of plant functional types (PFTs) for subarctic vegetation bio-types increase estimates of ozone-induced reduction of total biomass by $2.5$ to $17.4\,\unit{\%}$ in this flux-metric. Our study, suggests that assessments that use generic parameterizations are likely to underestimate the risk of \chem{O_3} damage in this region. We conclude that appropriate parameterization for the growing season and plant physiological responses to light and temperature are particularly important for accurate modeling of seasonal gas exchange. Efforts should be targeted towards accurately defining subarctic bio-type physiological response to climatic variables to improve regional and global scale biogeochemical cycling under current and future climates.
\end{abstract}

@@

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for ground-level ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the low-pressure systems for several consecutive months. In the period May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. While southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whole had close to normal precipitation, but with local variations.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and VOCs ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.
A total of 2079 forest fires of different sizes were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) \citep{SOU2019}. Coincident peak \chem{[O_3]} is found in ozone monitoring data from Svanhovd (Fig.~\ref{fig:data_svanvik_2018_2019} but also over whole northern Fennoscandia in July.
During the 2003 drought period, elevated ozone in Europe was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions \citep{JGR:Solberg2018}. It is likely that these factors also contributed to enhanced \chem{[O_3]} in 2018.\\

The described methods for risk assessment rely on accurate representation of plant physiological response to key environmental variables and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. It is important to understand the implications of using these more generic parameterizations for more extreme climate regions. We explore how risk estimates might be over or underestimated using generic vs bespoke subarctic parameterization. We also explore the role of \chem{[O_3]} variability and changing climate variables to understand how this risk might alter in the future with the advent of climate change. To this end, conditions at Svanhovd for the growing seasons 2018/19 shall serve as a reference for present and probable future conditions in northern Fennoscandia.
This allows an understanding of how the threat from ozone may change in the future and which are the key aspects of plant physiology that might determine the potential risk. Hence, we determine which parameters require special attention to define accurately for pollution impact assessment and more generally for describing gas exchange response to a changing climate that will influence biochemical cycling. This might also be one key to solve long-standing issues of earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface models \citep{GMD:Poulter2015,JAMES:Lawrence2019}.\\

@@

The foundation of the NIBIO Environment Centre Svanhovd dates back to the 1930s \citep{NIBIO_Svanhovd}. The NIBIO centre Svanhovd is located south of the settlement of Svanvik, where the Pasvik river, which is the border to Russia, lies only a few hundred meters to the east. It focuses on the research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

To relate local ozone concentration (\chem{[O_3]}) with visible damage on ozone sensitive plants, both an ozone bio-monitor (referred to as ozone garden) and a conventional ozone monitor using the existing atmospheric monitoring infrastructure had been installed during the 2018/19 growing season at the NIBIO centre. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). This ozone garden was likely the northern-most to date {\bf TODO: Reference?}. An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we qualitatively observed visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

@@

%\subsubsection{Ground-level ozone observations}
%\label{subsubsec:ebas}
\begin{figure}[t]
  %\includegraphics[width=8.3cm]{station_map_fennoscandia}
  \includegraphics[width=8.3cm]{fig03}
  \caption{Cap of the North. Locations of past and present (end of 2018) ozone observation sites in northern Fennoscandia were used in this study. For more details see Table~\ref{tab:ebas_obs}. The same color coding for each station as displayed herein will be used in all the following figures.}
  \label{fig:station_map_fennoscandia}
\end{figure}

A comprehensive overview over all past and present ozone observation sites in northern Fennoscandia is given in Table~\ref{tab:ebas_obs} and Fig.~\ref{fig:station_map_fennoscandia}, respectively. All stations with available long-term observations are located at higher altitude than Svanhovd. In the absence of local precursors, the stratification of the atmosphere usually leads to a higher abundances in ozone at higher altitude \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, perennial grassland, and heath. However, on closer inspection the vegetation differs substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi, $100\,\unit{m}$ above treeline (pine forest), and the surrounding vegetation mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass adhering to the \citet{WMOGuide2018} standards (Fig.\ref{fig:svanhovd_research_station}). In conjunction with the location at lower altitude, this results in generally lower \chem{[O_3]} at Svanhovd compared to Pallas.

Data from before the 1990s are available solely from Svanvik and Jergul. These data, however, do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably leads to drifts in the observed data and may impose a false trend. In 1997, the ozone monitoring at Jergul was relocated downstream of the river Karasjokka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a baseline shift in the combined data series, denoted as Jergul/Karasjok, which would be particularly problematic if these were the sole data for comparison. Only a few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are listed for completeness. There is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik (Supplement~Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}).

\begin{table}[t]
  \caption{Past and present ozone observation sites in northern Fennoscandia. Data available on \href{http://ebas.nilu.no/}{EBAS}.}
  \label{tab:ebas_obs}
  \begin{tabular}{lllllrl}
    \tophline
    Name       & Country & ID      & \multicolumn{3}{c}{Location} & Operational\\
    &         &         & lat             & lon               & alt            &\\
    &         &         & (\unit{^\circ N}) & (\unit{^\circ E})  & (\unit{m})     &\\
    \middlehline
    Esrange    & SWE     & SE0013R & 67.83           & 21.07             & 475            & 1991 -- 2018$^1$\\
    Janiskoski & RUS     & RU0001R & 68.93           & 28.85             & 118            & 1995 -- 1997\\
    Jergul     & NOR     & NO0030R & 69.45           & 24.60             & 255            & 1997 -- 2011\\
    Karasjok   & NOR     & NO0055R & 69.467          & 25.217            & 333            & 1988 -- 1997\\
    Pallas     & FIN     & FI0096G & 67.97           & 24.12             & 565            & 1995 -- present$^2$\\
    Svanvik    & NOR     & NO0047R & 69.45           & 30.03             & 30             & 1986 -- 1996$^3$\\
    \bottomhline
  \end{tabular}
  \belowtable{$^1$ Data available on EBAS until the end of 2018.\\
    $^2$ Data available on EBAS until the end of 2019.\\
    $^3$ Exclusive ozone monitoring in growing seasons 2018/19 for the present study operated by NILU.} % Table Footnotes
\end{table}

@@

\subsubsection{Ozone}
\label{subsubsec:anomal_ozone}
To assess the significance of 2018/19 anomalies, we estimate the bias due to the change in the tropospheric background \chem{[O_3]} since the 1990s and account for the missing data in June 2018. To reconstruct these missing data, we performed a Reynolds decomposition (Eq.~(\ref{eq:reynolds_decomp})) and cross-calibrate with ozone observations at Esrange and Pallas (companion paper). In the following, we will describe the bias-correction and discuss the significance of \chem{[O_3]} enhancements in 2018.

In 2018, the probability density functions (PDFs) of ozone concentration anomalies (\chem{\Delta[O_3]}) for Esrange and Pallas deviate clearly, but not significantly from the climatological mean (Supplement~Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals}). On average, \chem{[O_3]} had been elevated by $(1.9\pm 5.7)\,\unit{ppb}$ and $(1.9\pm 5.4)\,\unit{ppb}$, respectively. Both distributions are slightly skewed towards positive deviations with $\frac{1}{4}$ of days (3. quantile, q3) falling above $5.8\,\unit{ppb}$ and $5.6\,\unit{ppb}$, respectively.

@@

\section{DO3SE modeling}
\label{sec:do3se}
In the following, we investigate the difference between 2018 and 2019 concerning stomatal ozone uptake and assess which of the environmental factors might be most influential in determining \chem{O_3} damage risk. We model ozone uptake of different natural and semi-natural vegetation using the $\mathrm{DO_3SE}$ model and develop bespoke subarctic parameterizations for common plant species. First, we give an account of the methodology for deriving bespoke parameters, after which we present and discuss the modeling results. The $\mathrm{DO_3SE}$ model and conducted gap-filling methods are described in Appendix~\ref{appendix:do3se_model}.

@@

We deduced the dominant species on-site at the NIBIO Environment Centre Svanhovd and from Fig.~\ref{fig:svanhovd_research_station} and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Default parameters for these species or PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as \emph{mapping manual} abbreviated as MM in the following. We base our parameterization of coniferous trees on the MM's Boreal Norway spruce and deciduous trees are derived from MM's silver birch. For a comprehensive list of model parameters consult Supplement~Table~S1.

Initial test simulations with the MM PFTs revealed an unrealistically low stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 particularly for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance occured only during an extended warm period in late July. In ecological terms, this means that there was almost no growth of grass in the summer of 2019 -- a prediction which is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low, being the limiting factor of stomatal conductance in perennial grassland. Therefore, we saw the necessarity to propose bespoke parameterizations describing an acclimation of these species to the climatic conditions in the target region. Note, however, that these parameterizations represent hypothetical PFTs or species that have yet to be verified by experiments.

We hypothesize that perennial vegetation will likely prioritize their growth or fast adolescence and hence their carbon sequestration, during the short, subarctic growing season. Because the mapping manual version of the $\mathrm{DO_3SE}$ model currently does not simulate net photosynthesis ($A_\mathrm{n}$), we assume a first order proportionality between $A_\mathrm{n}$ and $g_\mathrm{sto}$ \citep{GCB:Medlyn2011} and effectively tune the temperature and light response functions ($f_\mathrm{T}$ and $f_\mathrm{light}$) for higher $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (see Appendix~\ref{appendix:do3se_model} for details). We further assume that $f_\mathrm{VPD}$ and $f_\mathrm{SWP}$ suffice our biomes. The procedure is graphically illustrated in Fig.~\ref{fig:f_temp_grassland} for grassland. We aim to increase the enclosed area between probability density function (PDF) and the respective response function. We show two climatological reference periods 1990s and 2000s to indicate that the vegetation is subject to ongoing climate change and hence continuous acclimation. Resulting response functions for coniferous and deciduous trees are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce}--\ref{fig:f_temp_birch}, respectively.

We define two different hypothetical temperature acclimation: \emph{subarctic} and \emph{cold}. For either, we assume that the hypothetical species are more tolerant to cold conditions and shift $f_\mathrm{T}$ towards colder temperatures accordingly. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. In the same way, \emph{subarctic} refers to a hypothetical species which is very tolerant to cold but is sensitive to high temperatures, and most efficient at cool temperatures ($T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000). The most extreme of the two acclimation types is \emph{subarctic}. For perennial grassland the resulting response functions are depicted together with the temperature PDF in Fig.~\ref{fig:f_temp_grassland}a).

Norway spruce was experimentally found to be active already at rather low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. Based on the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland \citep{TB:Kolari2007}, we estimate the optimal temperature interval $10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$. We assume a similar temperature acclimation for deciduous trees. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.

>>>>

%\title{The 2018 heatwave and its implications on ozone induced damage on vegetation in a subarctic climate}
\title{Characterizing subarctic European vegetation for land surface modeling of ozone pollution and climate risk}

@@

\Author[1,4]{Stefanie}{Falk}
\Author[2]{Ane V.}{Vollsnes}
\Author[2]{Aud B.}{Eriksen}
\Author[3]{Lisa}{Emberson}
\Author[3]{Connie}{O'Neill}
\Author[1]{Frode}{Stordal}
\Author[1]{Terje}{Koren Berntsen}

\affil[1]{Department of Geosciences, University of Oslo, Oslo, Norway}
\affil[2]{Department of Biosciences, University of Oslo, Oslo, Norway}
\affil[3]{Department of Environment and Geography, University of York, UK}
\affil[4]{Department of Geography, Ludwig Maximilians Universität, Munich, Germany}

@@

\begin{abstract}
  The unique vegetation of the subarctic region acclimatized to extremes of cold and midnight sun are likely to be at threat from the combined impacts of climate change and increasing ozone concentrations (\chem{[O_3]}). The atmospheric and climatic characteristics of the subarctic are known to lead to pronounced peak \chem{[O_3]} in spring. To date, only a few studies have been conducted to assess the response of subarctic vegetation to variations in climate and air pollution. This study looks to fill this knowledge gap by examining essential climate variables, in particular \chem{[O_3]}, over the past few decades. These data are used to assess the extent to which two recent years (2018 and 2019) deviate from climatic and \chem{[O_3]} norms and how these potentially more frequent future deviations may influence ozone damage to subarctic vegetation. We find that 2018 was an anomalously warm and bright year (in particular in spring and early summer) with higher than average \chem{O_3} concentrations in April/May and a high number of episodic peak \chem{O_3} concentration above $40\,\unit{ppb}$ in June--August, the latter in part attributable to forest fires in the Northern Hemisphere as well as warmer and sunnier conditions. We apply the integrated flux-metric Phytotoxic Ozone Dose (POD) to determine ozone risk and damage to vegetation as a function of \chem{[O_3]}, climate variables, and species-specific physiology. We find that bespoke parameterizations of plant functional types (PFTs) for subarctic vegetation bio-types result in a ozone-induced biomass loss of $2.5$ to $17.4\,\unit{\%}$. Our study, suggests that assessments that use generic parameterizations are likely to underestimate the risk of ozone damage in this region. We conclude that appropriate parameterization for the growing season and plant physiological responses to light and temperature are particularly important for accurate modeling of seasonal gas exchange. Efforts should be targeted towards accurately defining subarctic ecotypes' physiological response to essential climate variables to improve regional and global scale biogeochemical cycling under current and future climates.
\end{abstract}

@@

According to a report by the Norwegian Meteorological institute \citep{MetNOR2019}, the summer of 2018 was the warmest and driest ever recorded in eastern, western, and southern Norway. In the north (including Finnmark), it was amongst the warmest on record -- favorable conditions for ground-level ozone formation.
An unusually weak and northward shifted jet stream allowed for a persistent high-pressure system above northern Europe, including Fennoscandia which blocked the low-pressure systems for several consecutive months. In the period May -- July, southern Norway had temperatures $4\,\unit{^\circ C}$ above normal. While southern Norway had only about $60\,\unit{\%}$ of normal precipitation, northern Norway as a whole had close to normal precipitation, but with local variations.
Thermal stress on vegetation was exceptional not only in large parts of Fennoscandia but also in much of Europe, where the influence of the high-pressure system extended even over five months (April/May, July--September). These conditions gave rise to massive forest fires in different parts of Europe and thus an increase in ozone precursors. Boreal wildfires emit in addition to \chem{CO_2} also \chem{CO} ($\chem{[CO]}/\chem{[CO_2]} \propto 6-13\,\unit{\%}$) and VOCs ($\chem{[VOCs]}/\chem{[CO_2]} \propto 0.5-1.5\,\unit{\%}$) \citep{AE:Cofer1990}.
A total of 2079 forest fires of different sizes were registered in Norway in 2018, twice as many as in the preceding years 2016/17 \citep{DSB2019}, last accessed April 2020). In Sweden, about 500 fires had been reported (five times more than in a usual summer), and an estimated total of $25000\,\unit{hectare}$ burned down in central Sweden (G\"{a}vleborgs, J\"{a}mtlands, and Dalarnas l\"{a}n) \citep{SOU2019}. Coincident peak \chem{[O_3]} is found in ozone monitoring data from Svanhovd (Fig.~\ref{fig:data_svanvik_2018_2019} but also over whole northern Fennoscandia in July. Therefore conditions at Svanhovd for the growing seasons 2018/19 shall serve as a reference for present and probable future conditions in northern Fennoscandia in our study.

During the 2003 drought period, elevated ozone in Europe was promoted by a combination of various factors such as wildfires, reduced cloud cover (increased solar radiation), reduced dry deposition and turbulent mixing due to the stagnant weather conditions, and increased BVOC emissions \citep{JGR:Solberg2018}. It is likely that these factors also contributed to enhanced \chem{[O_3]} in 2018.\\

The described methods for risk assessment rely on accurate representation of plant physiological response to key environmental variables and are most often parameterized using data for less extreme climates, i.e., temperate or continental climates; and to a lesser extent, Boreal and Mediterranean climates. Parameterization of these models for more extreme bio-geographical regions such as subarctic climates has not been performed and relies on assuming that less extreme Boreal parameterization will be applicable. It is important to understand the implications of using these more generic parameterizations for more extreme climate regions. We explore how risk estimates might be over or underestimated using generic vs bespoke subarctic parameterization. We also explore the effects of the interannual variability of \chem{[O_3]} and essential climate variables (sensu World Meteorological Organization) on future risk of vegetation damage.
This allows an understanding of how the threat from ozone may change in the future and which are the key aspects of plant physiology that might determine the potential risk. Hence, we determine which parameters require special attention to define accurately for pollution impact assessment and more generally for describing gas exchange response to a changing climate that will influence biochemical cycling. This might also be one key to solve long-standing issues of earth system models (ESM) which often lack PFTs specifically representing subarctic conditions in their land surface models \citep{GMD:Poulter2015,JAMES:Lawrence2019}.\\

@@

The NIBIO centre Svanhovd focuses on the research of the special environment of the region, but also comprises long-term observations of various agrometeorological variables, i.e. $2\,\unit{m}$ temperature, precipitation, global irradiance, and soil temperature at different depths.

To relate local ozone concentration (\chem{[O_3]}) with visible damage on ozone sensitive plants, both an ozone bio-monitor (referred to as ozone garden) and a conventional ozone monitor using the existing atmospheric monitoring infrastructure had been installed during the 2018/19 growing season at the NIBIO centre. The locations of the atmospheric monitoring site and the ozone garden are marked in the aerial photography shown in Fig.~\ref{fig:svanhovd_research_station}a). An ozone garden consists of selected plant species that are sensitive to ozone and are likely to display visible injuries. Cultivated species were, e.g. clover, tobacco, and potatoes. As shown in Fig.~\ref{fig:svanhovd_research_station}b), the plants had to be protected from herbivores with a wire-mesh fence. In 2018, we qualitatively observed visible ozone damage on semi-natural vegetation (clover) and crops (tobacco). In contrast, no such visible damage was found on the clovers in 2019, although it was observed on the sensitive tobacco cultivar. This indicates that the vegetation may have been more affected by ozone in 2018 than in 2019.

@@

%\subsubsection{Ground-level ozone observations}
%\label{subsubsec:ebas}
\begin{figure}[t]
  %\includegraphics[width=8.3cm]{station_map_fennoscandia}
  \includegraphics[width=8.3cm]{fig03}
  \caption{Cap of the North. Locations of past and present (end of 2018) ozone observation sites in northern Fennoscandia used in this study. For more details see Table~\ref{tab:ebas_obs}. The same color coding for each station as displayed herein will be used in all the following figures. Figure adapted from \citet{ACP:Falk2021}. \textbf{TODO: Add map copy right}}
  \label{fig:station_map_fennoscandia}
\end{figure}

A comprehensive overview over all past and present ozone observation sites in northern Fennoscandia is given in Table~\ref{tab:ebas_obs} and Fig.~\ref{fig:station_map_fennoscandia}, respectively. All stations with available long-term observations are located at higher altitude than Svanhovd. In the absence of local precursors, the stratification of the atmosphere usually leads to a higher abundances in ozone at higher altitude \citep[e.g.,][]{AB:Klingberg2009}. The periphery vegetation at all sites is similar and consists mainly of pine forests, birch, perennial grassland, and heath. However, on closer inspection the vegetation differs substantially. Ozone measurements at Pallas are taken on top of mount Sammaltunturi, $100\,\unit{m}$ above treeline (pine forest), and the surrounding vegetation mainly consists of low vascular plants, mosses, and lichen \citep{BER:Hatakka2003}. At Svanhovd, the ozone measurements are taken amidst a meadow surrounded by agriculturally managed herbaceous plants and grass adhering to the \citet{WMOGuide2018} standards (Fig.\ref{fig:svanhovd_research_station}).

Data from before the 1990s are available solely from Svanvik and Jergul. These data do not follow the high-quality standard procedures implemented nowadays and have to be treated with care \citep{NILU2003}. Especially, ozone monitors did not undergo regular re-calibration. This probably lead to drifts in the observed data and may impose a false trend. In 1997, the ozone monitoring at Jergul was relocated downstream of the river Karasjohka to a site close by Karasjok. Although the same equipment has been used, this might have introduced a baseline shift in the combined data series, denoted as Jergul/Karasjok, which would be particularly problematic if these were the sole data for comparison. Only a few data have been taken upstream the Pasvik river at Janiskovski, hence these will not be taken into further consideration, but are listed for completeness. There is an overlap of about $4\,\unit{years}$ between data taken at Esrange, Jergul/Karasjok, and Svanvik (Supplement~Fig.~\ref{fig:ozone_timesseries_fenoscandic_obs}).

\begin{table}[t]
  \caption{Past and present ozone observation sites in northern Fennoscandia. Data available on \href{http://ebas.nilu.no/}{EBAS}. Table adapted from \citet{ACP:Falk2021}}
  \label{tab:ebas_obs}
  \begin{tabular}{lllllrl}
    \tophline
    Name       & Country & ID      & \multicolumn{3}{c}{Location} & Operational\\
    &         &         & lat             & lon               & alt            &\\
    &         &         & (\unit{^\circ N}) & (\unit{^\circ E})  & (\unit{m})     &\\
    \middlehline
    Esrange    & SWE     & SE0013R & 67.83           & 21.07             & 475            & 1991 -- 2018$^1$\\
    Janiskoski & RUS     & RU0001R & 68.93           & 28.85             & 118            & 1995 -- 1997\\
    Jergul     & NOR     & NO0030R & 69.45           & 24.60             & 255            & 1997 -- 2011\\
    Karasjok   & NOR     & NO0055R & 69.467          & 25.217            & 333            & 1988 -- 1997\\
    Pallas     & FIN     & FI0096G & 67.97           & 24.12             & 565            & 1995 -- present$^2$\\
    Svanvik    & NOR     & NO0047R & 69.45           & 30.03             & 30             & 1986 -- 1996$^3$\\
    \bottomhline
  \end{tabular}
  \belowtable{$^1$ Data available on EBAS until the end of 2018.\\
    $^2$ Data available on EBAS until the end of 2019.\\
    $^3$ Exclusive ozone monitoring in growing seasons 2018/19 for the present study operated by NILU.} % Table Footnotes
\end{table}

@@

\subsubsection{Ozone}
\label{subsubsec:anomal_ozone}
\textbf{TODO: Remove all content which is part of the technical note}
To assess the significance of 2018/19 anomalies, we estimate the bias due to the change in the tropospheric background \chem{[O_3]} since the 1990s and account for the missing data in June 2018. To reconstruct these missing data, we performed a Reynolds decomposition (Eq.~(\ref{eq:reynolds_decomp})) and cross-calibrate with ozone observations at Esrange and Pallas (companion paper). In the following, we will describe the bias-correction and discuss the significance of \chem{[O_3]} enhancements in 2018.

In 2018, the probability density functions (PDFs) of ozone concentration anomalies (\chem{\Delta[O_3]}) for Esrange and Pallas deviate clearly, but not significantly from the climatological mean (Supplement~Fig.~\ref{fig:ozone_climatology_fenoscandic_obs_residuals}). On average, \chem{[O_3]} had been elevated by $(1.9\pm 5.7)\,\unit{ppb}$ and $(1.9\pm 5.4)\,\unit{ppb}$, respectively. Both resemble normal distributions.

@@

\section{DO3SE modeling}
\label{sec:do3se}
In the following, we investigate the difference between 2018 and 2019 concerning stomatal ozone uptake and assess which of the environmental factors might be most influential in determining \chem{O_3} damage risk. We model ozone uptake of natural and semi-natural vegetation using the $\mathrm{DO_3SE}$ model and develop bespoke subarctic parameterizations for common plant species. First, we give an account of the methodology for deriving bespoke parameters, after which we present and discuss the modeling results. The $\mathrm{DO_3SE}$ model and conducted gap-filling methods are described in Appendix~\ref{appendix:do3se_model}.

@@

We deduced the dominant species on-site at the NIBIO Environment Centre Svanhovd and from Fig.~\ref{fig:svanhovd_research_station} and found perennial grassland, birch (generalized as deciduous trees), and Scots pine (generalized as coniferous trees). Default parameters for these species or PFTs are derived from \citet{EP:Simpson2007,GCB:Mills2011,ICP:MappingManual2017} and will be referred to as \emph{mapping manual} (MM) parameterizations in the following. We base our parameterization of coniferous trees on the MM's Boreal Norway spruce and deciduous trees on the MM's silver birch. For a comprehensive list of model parameters, consult Supplement~Table~S1.

Initial test simulations with the MM PFTs revealed an unrealistically low stomatal conductance at leaf-level ($G_\mathrm{sto}^\mathrm{leaf}$) in 2019 particularly for perennial grassland (Appendix~Fig.~\ref{fig:pody_mm_composit}). Substantial stomatal conductance occured only during an extended warm period in late July. In ecological terms, this means that there was almost no growth of grass in the summer of 2019 -- a prediction which is easily falsified by reality. We identified $f_\mathrm{T}$ as abnormally low, being the limiting factor of stomatal conductance in perennial grassland. Therefore, we saw the necessity to propose bespoke parameterizations describing an acclimation of of the PFTs to the climatic conditions in the target region. Note, however, that these parameterizations are hypothetical and have yet to be verified by experiments.

We assume that perennial vegetation will likely maximize carbon acquisition during the short subarctic growing season. Because the mapping manual version of the $\mathrm{DO_3SE}$ model currently does not simulate net photosynthesis ($A_\mathrm{n}$), we assume a first order proportionality between $A_\mathrm{n}$ and $g_\mathrm{sto}$ \citep{GCB:Medlyn2011} and effectively tune the temperature and light response functions ($f_\mathrm{T}$ and $f_\mathrm{light}$) for higher $\left<g_\mathrm{sto}\right>/g_\mathrm{max}$ (see Appendix~\ref{appendix:do3se_model} for details) to acclimate the growth potential to the local climate. We further assume that $f_\mathrm{VPD}$ and $f_\mathrm{SWP}$ suit our biomes. The procedure is graphically illustrated in Fig.~\ref{fig:f_temp_grassland} for grassland. We aim to increase the overlapping area between the probability density function (PDF) and the respective response function. We show two climatological reference periods 1990s and 2000s to indicate that the vegetation is subject to ongoing climate change and hence continuous acclimation. The resulting response functions for coniferous and deciduous trees are displayed in Appendix~Figs.~\ref{fig:f_temp_spruce}--\ref{fig:f_temp_birch}, respectively.

We define two different hypothetical temperature acclimations: \emph{subarctic} and \emph{cold}. For either, we assume that the hypothetical species are more tolerant to cold conditions and shift $f_\mathrm{T}$ towards colder temperatures accordingly. We construct \emph{cold} as representative for a species that is more tolerant to cold temperatures, but slightly less efficient at warm temperatures compared to MM. In the same way, \emph{subarctic} refers to a hypothetical species which is very tolerant to cold but is sensitive to high temperatures, and most efficient at cool temperatures ($T_\mathrm{opt}$ close to the climatological mean temperature of 1992--2000). The most extreme of the two acclimation types is \emph{subarctic}. For perennial grassland the resulting response functions are depicted together with a five month growing season temperature PDF of the two periods 1992-2000 and 2011-2019 in Fig.~\ref{fig:f_temp_grassland}a).

Norway spruce was experimentally found to be active already at rather low air temperatures and can reach $60\,\unit{\%}$ photosynthetic activity as early as \unit{doy}~100 \citep{TB:Kolari2007, TP:Wallin2013}. Based on the time series of \chem{CO_2} uptake and temperature at observation sites in southern ($\rightarrow$ \emph{cold}) and northern ($\rightarrow$ \emph{subarctic}) Finland \citep{TB:Kolari2007}, we estimate the optimal temperature interval $10\,\unit{^\circ C}\le T_\mathrm{opt} \le 15\,\unit{^\circ C}$. We assume similar temperature acclimations for deciduous trees. All bespoke temperature parameters are listed in Table~\ref{tab:sensitivity_tests_temp}.

With respect to light conditions, we assume that the light threshold for a $50\,\unit{\%}$ opening of plants' stomata may differ in subarctic species compared to species in less extreme climates. Analytically, we derive the inverse function $f_{\mathrm{light},k}^{-1}$ of Eq.~(\ref{eq:flight}) for each PFT or species $k$
\begin{equation}
  \gamma_k(f_\mathrm{light}) := f_{\mathrm{light}, k}^{-1} = -\frac{\ln(1-f_\mathrm{light})}{\alpha_k}.
  \label{eq:inverse_function}
\end{equation}
First, we solve Eq.~(\ref{eq:inverse_function}) for the MM default value $\alpha_\mathrm{MM}^k$ for each PFT or species $k$ at $0.5$ 
\begin{equation}
  \gamma^k(0.5) = -\frac{\ln(0.5)}{\alpha_\mathrm{MM}^k},
  \label{eq:inverse_function_halfway}
\end{equation}
and define a variation $\gamma' = \gamma \cdot \eta$ with $\eta \in \{0.8, 1.2\}$. We invert Eq.~(\ref{eq:inverse_function}) for $\alpha$ and find $\alpha'(\gamma')$. The resulting functions for perennial grassland are shown in Fig.~\ref{fig:f_temp_grassland}b) (for deciduous and coniferous trees refer Appendix~Figs.~\ref{fig:f_temp_spruce}b)--\ref{fig:f_temp_birch}b)). All derived parameters are tabulated in Table~\ref{tab:sensitivity_tests_light}.
